---
format: gfm
execute: 
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(sf)
library(patchwork)
```


```{r}
f = list.files(pattern = "xlsx")
f_last = tail(f, 1)
school_ms = readxl::read_excel(f_last)
nms = names(school_ms)
#  [1] "Local Authority"       "LA Name"               "Site"                 
#  [4] "URN"                   "Current Accreditation" "Survey Type"          
#  [7] "School Year"           "Survey Month"          "Survey Year"          
# [10] "Participation %"       "Total Actual"          "Walk"                 
# [13] "Cycle"                 "Scoot/Skate"           "Public Bus"           
# [16] "School Bus"            "P&S"                   "Train"                
# [19] "Car Share"             "Car"                   "Motorcyle"            
# [22] "% Walk"                "% Cycle"               "% Scoot/Skate"        
# [25] "% Public Bus"          "% School Bus"          "% P&S"                
# [28] "% Train"               "% Car Share"           "% Car"                
# [31] "% Motorcyle"
accred = table(school_ms$`Current Accreditation`)
#    Approved TP   Excellent TP        Good TP Outstanding TP   Very Good TP 
#           7687           2488          17001           2720           4020 

```

```{r}

```

A plot of the counts of students using different modes of travel to school over time (2018/2019 to 2022/2023 shown as years for which we have a high response rate), with facets showing different travel plan (TP) ratings, is shown below.

```{r}
#| label: mode-count-time
school_ms_long = school_ms |>
  select(`LA Name`:Motorcyle) |>
  select(-Site, -`Survey Year`) |>
  pivot_longer(cols = nms[12:21], names_to = "Mode", values_to = "count")
# Create simplified mode breakdown:
school_ms_long = school_ms_long |>
  mutate(
    `Mode (full)` = Mode,
    Mode = case_when(
      Mode %in% c("Walk", "Scoot/Skate") ~ "Walk/Scoot",
      Mode %in% "Cycle" ~ "Cycle",
      Mode %in% c("Public Bus", "School Bus", "P&S", "Train") ~ "Public Transport",
      Mode %in% c("Car Share", "Car", "Motorcyle") ~ "Private Transport",
      TRUE ~ "Other"
    )
  ) |>
  # Ordered factor:
  mutate(Mode = factor(Mode, levels = c("Walk/Scoot", "Cycle", "Public Transport", "Private Transport", "Other")))
```

The modeshift dataset contains a detailed breakdown by mode, with the number of observations for each mode and year shown below:

```{r}
#| label: mode-full
school_ms_long |>
  group_by(`School Year`, Mode = `Mode (full)`) |>
    summarise(count = sum(count, na.rm = TRUE)) |>
    ungroup() |>
  ggplot(aes(x = `School Year`, y = count, fill = Mode), group = Mode) +
  geom_bar(stat = "identity") +
    facet_wrap(~Mode) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
school_msy = school_ms_long |>
  group_by(`School Year`, `Current Accreditation`, Mode) |>
    summarise(count = sum(count, na.rm = TRUE)) |>
    ungroup() 
# table(school_msy$`School Year`)
# summary(school_msy)

school_msy |>
  filter(`School Year` >= "2018/2019" & `School Year` <= "2022/2023") |>
  ggplot(aes(x = `School Year`, y = count, fill = Mode), group = Mode) +
  geom_bar(stat = "identity") +
    facet_wrap(~`Current Accreditation`) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

school_ids_subset = school_ms_long |>
  group_by(URN, `School Year`) |>
  mutate(n_responses = sum(count, na.rm = TRUE)) |>
  filter(n_responses > 0) |>
  summarise(n = n()) |>
  mutate(school_years_char = paste(`School Year`, collapse = ", ")) |>
  filter(str_detect(school_years_char, "2018/2019, 2019/2020, 2020/2021, 2021/2022, 2022/2023")) |>
  pull(URN) |>
  unique()

```

To get a handle on the rate of change, we'll focus on a subset of schools that have continuous records for the 5 years from 2018/2019 to 2022/2023.
Of the `r length(unique(school_ms$URN))` schools in the dataset, `r length(school_ids_subset)` schools have continuous records for the 5 years from 2018/2019 to 2022/2023.


```{r}
#| label: mode-proportion-time-subset
school_proportions = school_ms_long |>
  group_by(`School Year`, Mode) |>
    summarise(count = sum(count, na.rm = TRUE)) |>
    # Convert to % for each year and accreditation:
    group_by(`School Year`) |>
    mutate(Proportion = count / sum(count, na.rm = TRUE)) |>
    ungroup() |>
  filter(`School Year` >= "2018/2019" & `School Year` <= "2022/2023")
school_proportions_subset = school_ms_long |>
  filter(URN %in% school_ids_subset) |>
  group_by(`School Year`, Mode) |>
    summarise(count = sum(count, na.rm = TRUE)) |>
    # Convert to % for each year and accreditation:
    group_by(`School Year`) |>
    mutate(Proportion = count / sum(count, na.rm = TRUE)) |>
    ungroup() |>
  filter(`School Year` >= "2018/2019" & `School Year` <= "2022/2023")
g1 = school_proportions |>
  ggplot(aes(x = `School Year`, y = Proportion, fill = Mode), group = Mode) +
  geom_bar(stat = "identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    # % y lab:
    scale_y_continuous(labels = scales::percent_format())
g2 = school_proportions_subset |>
  ggplot(aes(x = `School Year`, y = Proportion, fill = Mode), group = Mode) +
  geom_bar(stat = "identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    # % y lab:
    scale_y_continuous(labels = scales::percent_format())
# g1 | g2 # very similar
g2
```

Let's calculate the rate of change in proportions:

```{r}
school_proportions_urn = school_ms_long |>
  filter(URN %in% school_ids_subset) |>
  group_by(`School Year`, URN, Mode) |>
    summarise(count = sum(count, na.rm = TRUE)) |>
    # Convert to % for each year and accreditation:
    group_by(`School Year`, URN) |>
    mutate(Proportion = count / sum(count, na.rm = TRUE)) |>
    ungroup() 
# summary(school_proportions_urn)
```

```{r}
school_proportions_wide = school_proportions_urn |>
  select(-count) |>
  pivot_wider(names_from = Mode, values_from = Proportion)
```

```{r}
# Summarised dataset with % uptake for each active mode:
school_proportions_urn_summary = school_proportions_wide |>
  # Filter to the 5 years of interest:
  filter(`School Year` %in% c("2018/2019", "2019/2020", "2020/2021", "2021/2022", "2022/2023")) |>
  arrange(URN, `School Year`) |>
  group_by(URN) |>
  summarise(
    `Walk/Scoot` = `Walk/Scoot`[[5]] - `Walk/Scoot`[[1]],
    Cycle = Cycle[[5]] - Cycle[[1]],
    Active = `Walk/Scoot` + Cycle,
    `Public Transport` = `Public Transport`[[5]] - `Public Transport`[[1]],
    `Private Transport` = `Private Transport`[[5]] - `Private Transport`[[1]]
  ) |>
  # Classify schools by mode shift with breaks at -0.5, 0, 0.01, 0.02, 0.05, 0.08, 1
  mutate(
    Group = case_when(
      Active < 0 ~ "Decrease",
      Active < 0.02 ~ "Small Increase",
      Active < 0.05 ~ "Increase",
      Active < 0.08 ~ "Large Increase",
      TRUE ~ "Very Large Increase"
    )
  ) |>
  # As factor:
  mutate(Group = factor(Group, levels = c("Decrease", "Small Increase", "Increase", "Large Increase", "Very Large Increase"))
  ) 

# table(school_proportions_urn_summary$Group)

# summary(school_proportions_urn_summary)
school_proportions_urn_summary_long = school_proportions_urn_summary |>
  pivot_longer(cols = c(`Walk/Scoot`, Cycle, `Public Transport`, `Private Transport`), names_to = "Mode", values_to = "change")

```

Let's visualise the change in % active travel by group, with breaks at -0.5, 0, 0.01, 0.02, 0.05, 0.08, 1:

```{r}
#| label: mode-proportion-time-single
school_proportions_urn_summary_long |>
  filter(Mode != "Private Transport") |>
  ggplot(aes(x = Group, y = change, fill = Mode), group = Mode) +
  geom_boxplot() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    # % y lab:
    scale_y_continuous(labels = scales::percent_format())
```

The graph shows that schools that have increased active travel (excluding walking to public transport) have done so primarily by boosting walking.
However, some schools have been very successful in boosting cycling.
Summary statistics for the 10% of schools with the largest increase in active travel, for walking and cycling, are shown below:


```{r}
success_10pct_walk = school_proportions_urn_summary |>
  arrange(desc(`Walk/Scoot`)) |>
  slice(1:round(0.1 * n()))
success_10pct_cycle = school_proportions_urn_summary |>
  arrange(desc(Cycle)) |>
  slice(1:round(0.1 * n()))
# mean(success_10pct_walk$`Walk/Scoot`)
# mean(success_10pct_cycle$Cycle)
# # Change per year:
mean_change_walk = mean(success_10pct_walk$`Walk/Scoot`) / 5
mean_change_cycle = mean(success_10pct_cycle$Cycle) / 5  
```

Overall, we found that the 10% of schools with the largest increase in walking boosted walking by an average of `r round(mean_change_walk * 100, 1)` percentage points per year.
The 10% of schools with the largest increase in cycling boosted cycling by an average of `r round(mean_change_cycle * 100, 1)` percentage points per year.